News Across Languages - Cross-Lingual Document Similarity and Event Tracking

### 2015-11-10 rupnik2015.tex
minor revision

@@

In today's world, we follow news which is distributed globally. Significant events are reported by different sources and in different languages. In this work, we address the problem of tracking of events in a large multilingual stream. Within a recently developed system Event Registry  \shortcite{Leban2014W,Leban2014I} we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.  Taking a multilingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia. This allows us to compute the similarity of any two articles regardless of language. Building on previous work, we show there are methods which scale well and can compute a meaningful similarity between articles from languages with little or no direct overlap in the training data.
Using this capability, we then propose an approach to link clusters of articles across languages which represent the same event. We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality and robustness of the similarity measure and the linking algorithm.
\end{abstract}

@@

Content on the internet is becoming increasingly multilingual. A prime example is Wikipedia - in 2001, the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%. At the same time, online news has begun to dominate reporting of current events. However, machine translation remains relatively rudimentary - allowing people to understand simple phrases on web pages, but remain inadequate for more advanced understanding of text. In this paper we consider the intersection of these developments: how to track events which are reported about in multiple languages.

The term event is vague and ambiguous, but for the practical purposes, we define it as ``any significant happening that is being reported about in the media.'' Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{fig:event2}). Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event. Generally, events are more specific than general themes as the time component plays an important role -- for example, the two wars in Iraq would be considered as separate events.

@@

 We consider a high volume of articles in different languages. By using a language detector, the stream is split into separate monolingual streams. Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system \shortcite{Leban2014W,Leban2014I}. Our main goal in this paper is to connect such clusters (representations of events) across languages, that is, to detect that a set of articles in language $A$ reports on the same event as a set of articles in language $B$.

Our approach to linking clusters across languages combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and semantic annotation of documents, which enables an alternative way to comparing documents. Since this work represents a complicated pipeline, we concentrate on these two specific elements. Overall, the approach should be considered from a systems' perspective (considering the system as a whole) rather than considering these problems in isolation.

The first approach represents a continuation of previous work \shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.  The representations could be interpreted as multilingual topics, which were then used as proxies to compute cross-lingual similarities between documents. To learn the representations, we use Wikipedia as a training corpus. Significantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc. which have significant overlap in article coverage, but also smaller languages (in terms of number of Wikipedia articles) such as Slovenian and Hindi, which may have a negligible overlap in article coverage. We can then define a similarity between any two articles regardless of language,  which allows us to cluster the articles according to topic. The underlying assumption is that articles describing the same event are similar  and will therefore be put into the same cluster.

@@

The paper is organized as follows: we first provide an overview of the system as a whole in Section~\ref{sec:pipeline}, which includes a subsection that summarizes the main system requirements. We then present related work in Section~\ref{sec:related}. The related work covers work on cross-lingual document similarity as well as work on cross-lingual cluster linking. In Section~\ref{sec:crosslingual}, we introduce the problem of cross-lingual document similarity computation, which is followed by an overview  and description of the main cross-lingual similarity approaches we consider in detail, including a new approach based on hub languages.
%The approach in Section~\ref{sec:kmeans} serves as a baseline, and the approaches in Section~\ref{sec:CCA} and Section~\ref{sec:LSI} serve as the basis on which we proposed a new approach presented in Section~\ref{sec:hublang}
 In Section~\ref{sec:linking}, we introduce the central problem of cross-lingual linking of clusters of news articles and our approach that combines the cross-lingual similarity functions with knowledge extraction based techniques. Finally, we present and interpret the experimental results in Section~\ref{sec:evaluation} and discuss conclusions and point out several promising future directions.

@@

We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry \shortcite{Leban2014W,Leban2014I}. Event Registry is a repository of events, where events are automatically identified by analyzing news articles that are collected from numerous news outlets all over the world. The important components in the pipeline of the Event Registry are shown in Figure~\ref{fig:erpipeline}. We will now briefly describe the main components.

@@

Collected articles are first semantically annotated by identifying mentions of relevant concepts -- either entities or important keywords. The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base. The algorithm for semantic annotation is based on~\cite{zhang2014saaacamactat}. As a part of the semantic annotation we also analyze the dateline of the article to identify the location of the described event as well as extract dates mentioned in the article using a set of regular expressions. Since articles are frequently revised we also detect if the collected article is just a revision of a previous one so that we can use this information in next phases of the pipeline. The last important processing step on the document level is to efficiently identify which articles in other available languages are most similar to this article. The methodology for this task is one of the main contributions of this paper and is explained in details in Section~\ref{sec:crosslingual}).

@@

Our goal is to build a system that monitors global media and analyzes how events are being reported on. Our approach consists of two steps: tracking events separately in each language (based on language detection and an online clustering approach) and then connecting them. The pipeline must be able to process millions of articles per day and perform billions of similarity computations each day. Both step rely heavily on similarity computation, which implies that this must be highly scalable.

@@

\textbf{Translation and dictionary based}. The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity, see  \cite{multilingualBook}\cite{plagiarism} for several variations of translation based approaches. One can use free tools such as Moses \shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}. There are two issues with such approaches: they solve a harder problem than needs to be solved and they are less robust to training resource quality - large sets of translated sentences are typically needed. Training Moses for languages with scarce linguistic resources is thus problematic. The issue with using online services such as Google Translate is that the APIs are limited and not free. The operation efficiency and cost requirements make translation-based approaches less suited for our system. Closely related are works Cross-Lingual Vector Space Model (CL-VSM) \cite{plagiarism} and the approach presented in \cite{pouliquen2008story} which both compare documents by using dictionaries, which in both cases are EuroVoc dictionaries \cite{eurovoc}. The generality of such approaches is limited by the quality of available linguistic resources, which may be scarce or non-existent for certain language pairs.

\textbf{Probabilistic topic models}. There exist many variants to modelling documents in a language independent way by using probabilistic graphical models. The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA) \shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA) \cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)\cite{PCL_LSA} and Polylingual Topic Models (PLTM) \shortcite{polyLDA} which is a Bayesian version of PCLLSA. The methods (except for CPLSA) describe the multilingual document collections as samples from generative probabilistic models, with variations on the assumptions on the model structure. The topics represent latent variables that are used to generate observed variables (words), a process specific to each language. The parameter estimation is posed as an inference problem which is typically intractable and one usually solves it using approximate techniques. Most variants of solutions are based on Gibbs sampling or Variational Inference, which are nontrivial to implement and may require an experienced practitioner to apply. Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which must be solved.

\textbf{Matrix factorization}. Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI\shortcite{cl_lsi}\cite{multilingualBook}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}. The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes. In addition, OPCA forces the vocabulary sizes for all languages to be the same, which is less intuitive. For our setting, the method in \cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size). Our proposed approach combines CCA and CL-LSI. Another closely related method is Cross-Lingual Explicit Semantic Analysis (CL-ESA) \cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents. It can be interpreted as using the sample covariance matrix between features of two languages to define the dot product which is used to compute similarities.
The authors of CL-ESA compare it to CL-LSI and find that CL-LSI can outperform CL-ESA in an information retrieval, but is costlier to optimize over a large corpus (CL-ESA requires no training). We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to 10,000 document limit reported in \cite{ESA}). In addition, CL-ESA is less suited for computing similarities between two large monolingual streams. For example, each day we have to compute similarities between 500,000 English and 500,000 German news articles. Comparing each German news article with 500,000 English news articles is either prohibitively slow (involves projecting all English articles on Wikipedia) or consumes too much memory (involves storing the projected English articles, which for a Wikipedia of size 1,000,000 is a 500,000 by 1000,0000 non-sparse matrix).

\textbf{Monolingual}. Finally, related work includes monolingual approaches that treat document written in different languages in a monolingual fashion. The intuition is that named entities (for example, ``Obama'') and cognate words (for example, ``tsunami'') are written in the same or similar fashion in many languages. For example, the Cross-Language Character n-Gram Model (CL-CNG) \cite{plagiarism} represents documents as bags of character n-grams. Another approach is to use language dependent keyword lists based on cognate words \cite{pouliquen2008story}. These approaches may be suitable for comparing documents written in languages that share a writing system, which does not apply to the case of global news tracking.

@@

\section{Cross-lingual Document Similarity}
\label{sec:crosslingual}
Document similarity is an important component in techniques from text mining and natural language processing. Many techniques use the similarity as a black box, e.g., a kernel in Support Vector Machines. Comparison of documents (or other types of text snippets) in a monolingual setting is a well-studied problem in the field of information retrieval \cite{Salton88term-weightingapproaches}. We first formally introduce the problem followed by a description of  our approach.

@@

\noindent\textbf{Document representation.}
The standard vector space model \cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to a word or a phrase in a fixed vocabulary. Formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.

@@

\noindent\textbf {Mono-lingual similarity.}
A common way of computing similarity between documents is \emph{cosine similarity},
$$sim(d_1, d_2) = \frac{\langle \phi(d_1), \phi(d_2)\rangle}{\|\phi(d_1)\| \|\phi(d_2)\|},$$
where $\langle \cdot,\cdot \rangle$ and $\|\cdot\|$ are standard inner product and Euclidean norm. When dealing with two or more languages, one could ignore the language information
and build a vector space using the union of tokens over the languages. A cosine similarity function in such a space can be useful to some extent, for example "internet" or "Obama" may appear both in Spanish and English texts and the presence of such terms in both an English and a Spanish document would contribute to their similarity. In general however, large parts of vocabularies may not intersect, which means that given a language pair, many words in both languages cannot contribute to the similarity score, which makes the similarity function very insensitive to the data.

\noindent\textbf {Cross-lingual similarity.}
Processing a multilingual dataset results in several vector spaces with varying dimensionality, one for each language. The dimensionality of the vector space corresponding to the $i$-th language is denoted by $n_i$ and the vector space model mapping is denoted by $\phi_i : \text{text} \rightarrow \RR^{n_i}$.
The similarity between documents in language $i$ and language $j$ is defined as a bilinear operator represented as a matrix $S_{i,j} \in \RR^{n_i \times n_j}$:
$$sim_{i,j}(d_1, d_2) = \frac{ \langle \phi_i (d_1), S_{i,j} \phi_j (d_2) \rangle }{\|\phi_i(d_1)\| \|\phi_j(d_2)\|},$$
where $d_1$ and $d_2$ are documents written in the $i$-th and $j$-th language respectively. If the maximal singular value of $S_{i,j}$ is bounded by $1$, then the similarity scores will lie on the interval $[-1, 1]$. We will provide an overview of the models in Section \ref{sec:models} and then introduce additional notation in \ref{sec:notation}. Starting with Section \ref{sec:kmeans} and ending with Section \ref{sec:hublang} we will describe some approaches to computing $S_{i,j}$ given training data.

\subsection{Cross-Lingual Models}\label{sec:models}
In this section, we will describe several approaches to the problem of computing multilingual similarities introduced in Section~\ref{sec:tfidf}. We present four approaches:
a simple approach based on $k$-means clustering in Section~\ref{sec:kmeans}, a standard approach based on singular value decomposition in Section~\ref{sec:LSI}, a related
approach called Canonical Correlation Analysis (CCA) in Section~\ref{sec:CCA} and finally a new method, which is an extension of CCA to more than two languages in Section~\ref{sec:hublang}.
%
CCA can be used to find correlated patterns for a pair of languages, whereas the extended method optimizes a
Sum of Squared Correlations (SSCOR) between several language pairs, which was introduced in \cite{Kettenring}. The SSCOR problem is difficult to solve in our setting (hundreds of thousands of features, hundreds of thousands of examples). To tackle this, we propose a method which consists of two ingredients.
 The first one is based on an observation that certain datasets (such as Wikipedia) are biased towards one language (English for Wikipedia), which can be exploited
 to reformulate a difficult optimization problem as an eigenvector problem. The second ingredient is dimensionality reduction using CL-LSI, which
 makes the eigenvector problem computationally and numerically tractable.

@@

The cross-lingual similarity models presented in this paper are based on comparable corpora. A \emph{comparable corpus} is a collection of documents in multiple languages, with alignment between documents that are of the same topic, or even a rough translation of each other. Wikipedia is an example of a comparable corpus, where a specific entry can be described in multiple languages (e.g. ``Berlin" is currently described in 222 languages). News articles represent another example, where the same event can be described by newspapers in several languages.

More formally, a \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a tuple of $m$ documents on the same topic (comparable), where $u_i$ is the document written in language $i$. Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least \textbf{two nonempty documents}. This means that in our analysis we discard strictly monolingual documents for which no cross-lingual information is available. A comparable corpus $D = {d_1, \ldots, d_s}$ is a collection of $s$ multilingual documents. By using the vector space model, we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times s}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$. Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$ and the matrices respect the document alignment - the vector $X_i^\ell$ corresponds to the TFIDF vector of the $i$-th component of multilingual document $d_\ell$. We use $N$ to denote the total row dimension of $X$, i.e. $N:= \sum_{i=1}^m n_i$. See the Figure~\ref{fig:stacked_matrices} for an illustration of the introduced notation.

\begin{figure}[tbp]
\centering
\includegraphics[width=9cm]{stacked_matrices.pdf}
\caption{\label{fig:stacked_matrices} Multilingual corpora and their matrix representations using the vector space model.}
\end{figure}

We will now describe four models to cross-lingual similarity computation in the next sections.
\subsection{$k$-means}\label{sec:kmeans}

The $k$-means algorithm is perhaps the most well-known and widely-used clustering algorithm. Here, we present its application
to computing cross-lingual similarities. The idea is based on concatenating the corpus matrices, running standard $k$-means clustering to obtain the matrix of centroids, ``reversing" the concatenation step to obtain a set of aligned bases, which are finally used to compute cross-lingual similarities. See the Figure~\ref{fig:kmeans} for an illustration of process of computing multilingual clusters and using them to change the coordinates to cluster centroid ``topic" space.

@@

We then run the $k$-means algorithm \cite{kmeans} and obtain a centroid matrix $C \in \RR^{N \times k}$, where the $k$ columns represent centroid vectors. The centroid matrix can be split vertically into $m$ blocks: $$C = [C_1^T \cdots C_m^T]^T,$$ according to the number of dimensions of each language, i.e. $C_i \in \RR^{n_i \times k}$.
%
To reiterate, the matrices $C_i$ are computed using a multilingual corpus matrix $X$ (based on Wikipedia for example).

@@

The resulting matrix (when appropriately scaled to have unit norm) for similarity computation between language $i$ and language $j$ is defined as (see the reasoning in the next paragraph):
$$C_i(C_i^T C_i)^{-1} (C_j^T C_j)^{-1} C_j.$$

The matrix is a result of mapping documents in a language independent space using pseudo-inverses of the centroid matrices $P_i = (C_i^T C_i)^{-1} C_i$ and then comparing them using the standard inner product, which results in the matrix $P_i^T P_j$. For the sake of presentation, we assumed that the centroid vectors are linearly independent (an independent subspace could be obtained using an additional Gram-Schmidt step~\cite{golub} on the matrix $C$, if this was not the case).

In the next section we will describe our second approach.
\subsection{Cross-Lingual Latent Semantic Indexing}\label{sec:LSI}

>>>>

\makeatletter
\setlength{\@fptop}{0pt}
\makeatother

@@

In today's world, we follow news which is distributed globally. Significant events are reported by different sources and in different languages. In this work, we address the problem of tracking of events in a large multilingual stream. Within a recently developed system Event Registry~\shortcite{Leban2014W,Leban2014I} we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.  Taking a multilingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia. This allows us to compute the similarity of any two articles regardless of language. Building on previous work, we show there are methods which scale well and can compute a meaningful similarity between articles from languages with little or no direct overlap in the training data.
Using this capability, we then propose an approach to link clusters of articles across languages which represent the same event. We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality and robustness of the similarity measure and the linking algorithm.
\end{abstract}

@@

Content on the Internet is becoming increasingly multilingual. A prime example is Wikipedia. In 2001, the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%. At the same time, online news has begun to dominate reporting of current events. However, machine translation remains relatively rudimentary. It allows people to understand simple phrases on web pages, but remains inadequate for more advanced understanding of text. In this paper we consider the intersection of these developments: how to track events which are reported about in multiple languages.

The term event is vague and ambiguous, but for the practical purposes, we define it as ``any significant happening that is being reported about in the media.'' Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure~\ref{fig:event2}). Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event. Generally, events are more specific than general themes as the time component plays an important role -- for example, the two wars in Iraq would be considered as separate events.

@@

 We consider a high volume of articles in different languages. By using a language detector, the stream is split into separate monolingual streams. Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system~\shortcite{Leban2014W,Leban2014I}. Our main goal in this paper is to connect such clusters (representations of events) across languages, that is, to detect that a set of articles in language $A$ reports on the same event as a set of articles in language $B$.

Our approach to link clusters across languages combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and semantic annotation of documents, which enables an alternative way to comparing documents. Since this work represents a complicated pipeline, we concentrate on these two specific elements. Overall, the approach should be considered from a systems' perspective (considering the system as a whole) rather than considering these problems in isolation.

The first ingredient of our approach to link clusters across languages represents a continuation of previous work~\shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.  The representations could be interpreted as multilingual topics, which were then used as proxies to compute cross-lingual similarities between documents. To learn the representations, we use Wikipedia as a training corpus. Significantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc. which have significant overlap in article coverage, but also smaller languages (in terms of number of Wikipedia articles) such as Slovenian and Hindi, which may have a negligible overlap in article coverage. We can then define a similarity between any two articles regardless of language,  which allows us to cluster the articles according to topic. The underlying assumption is that articles describing the same event are similar  and will therefore be put into the same cluster.

@@

The paper is organized as follows: we first provide an overview of the system as a whole in Section~\ref{sec:pipeline}, which includes a subsection that summarizes the main system requirements. We then present related work in Section~\ref{sec:related}. The related work covers work on cross-lingual document similarity as well as work on cross-lingual cluster linking. In Section~\ref{sec:crosslingual}, we introduce the problem of cross-lingual document similarity computation and describe several approaches to the problem, most notably a new approach based on hub languages.
%The approach in Section~\ref{sec:kmeans} serves as a baseline, and the approaches in Section~\ref{sec:CCA} and Section~\ref{sec:LSI} serve as the basis on which we proposed a new approach presented in Section~\ref{sec:hublang}
 In Section~\ref{sec:linking}, we introduce the central problem of cross-lingual linking of clusters of news articles and our approach that combines the cross-lingual similarity functions with knowledge extraction based techniques. Finally, we present and interpret the experimental results in Section~\ref{sec:evaluation} and discuss conclusions and point out several promising future directions.

@@

We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry~\shortcite{Leban2014W,Leban2014I}. Event Registry is a repository of events, where events are automatically identified by analyzing news articles that are collected from numerous news outlets all over the world. The important components in the pipeline of the Event Registry are shown in Figure~\ref{fig:erpipeline}. We will now briefly describe the main components.

@@

Collected articles are first semantically annotated by identifying mentions of relevant concepts -- either entities or important keywords. The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base. The algorithm for semantic annotation uses machine learning to detect significant terms within unstructured text and link them to the appropriate Wikipedia articles. The approach models link probability and combines prior word sense distributions with context based sense distributions. The details are reported in~\cite{Milne:2008:LLW:1458082.1458150} and ~\cite{zhang2014saaacamactat}. As a part of the semantic annotation we also analyze the dateline of the article to identify the location of the described event as well as to extract dates mentioned in the article using a set of regular expressions. Since articles are frequently revised we also detect if the collected article is just a revision of a previous one so that we can use this information in next phases of the pipeline. The last important processing step on the document level is to efficiently identify which articles in other available languages are most similar to this article. The methodology for this task is one of the main contributions of this paper and is explained in details in Section~\ref{sec:crosslingual}.

@@

Our goal is to build a system that monitors global media and analyzes how events are being reported on. Our approach consists of two steps: tracking events separately in each language (based on language detection and an online clustering approach) and then connecting them. The pipeline must be able to process millions of articles per day and perform billions of similarity computations each day. Both steps rely heavily on similarity computation, which implies that this must be highly scalable.

@@

\textbf{Translation and dictionary based}. The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity, see  ~\cite{multilingualBook,plagiarism} for several variations of translation based approaches. One can use free tools such as Moses~\shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}. There are two issues with such approaches: they solve a harder problem than needs to be solved and they are less robust to training resource quality - large sets of translated sentences are typically needed. Training Moses for languages with scarce linguistic resources is thus problematic. The issue with using online services such as Google Translate is that the APIs are limited and not free. The operation efficiency and cost requirements make translation-based approaches less suited for our system. Closely related are works Cross-Lingual Vector Space Model (CL-VSM)~\cite{plagiarism} and the approach presented in~\cite{pouliquen2008story} which both compare documents by using dictionaries, which in both cases are EuroVoc dictionaries~\cite{eurovoc}. The generality of such approaches is limited by the quality of available linguistic resources, which may be scarce or non-existent for certain language pairs.

\textbf{Probabilistic topic models}. There exist many variants to modelling documents in a language independent way by using probabilistic graphical models. The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA)~\shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA)~\cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)~\cite{PCL_LSA} and Polylingual Topic Models (PLTM)~\shortcite{polyLDA} which is a Bayesian version of PCLLSA. The methods (except for CPLSA) describe the multilingual document collections as samples from generative probabilistic models, with variations on the assumptions on the model structure. The topics represent latent variables that are used to generate observed variables (words), a process specific to each language. The parameter estimation is posed as an inference problem which is typically intractable and one usually solves it using approximate techniques. Most variants of solutions are based on Gibbs sampling or Variational Inference, which are nontrivial to implement and may require an experienced practitioner to be applied. Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which must be solved.

\textbf{Matrix factorization}. Several matrix factorization based approaches exist in the literature. The models include: Non-negative matrix factorization based~\cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI~\shortcite{cl_lsi,multilingualBook}, Canonical Correlation Analysis (CCA)~\cite{Hotelling}, Oriented Principal Component Analysis (OPCA)~\cite{platt2010translingual}. The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes. In addition, OPCA forces the vocabulary sizes for all languages to be the same, which is less intuitive. For our setting, the method in ~\cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size). Our proposed approach combines CCA and CL-LSI. Another closely related method is Cross-Lingual Explicit Semantic Analysis (CL-ESA)~\cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents. It can be interpreted as using the sample covariance matrix between features of two languages to define the dot product which is used to compute similarities.
The authors of CL-ESA compare it to CL-LSI and find that CL-LSI can outperform CL-ESA in an information retrieval, but is costlier to optimize over a large corpus (CL-ESA requires no training). We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to the 10,000 document limit reported in~\cite{ESA}). In addition, CL-ESA is less suited for computing similarities between two large monolingual streams. For example, each day we have to compute similarities between 500,000 English and 500,000 German news articles. Comparing each German news article with 500,000 English news articles is either prohibitively slow (involves projecting all English articles on Wikipedia) or consumes too much memory (involves storing the projected English articles, which for a Wikipedia of size 1,000,000 is a 500,000 by 1000,0000 non-sparse matrix).

\textbf{Monolingual}. Finally, related work includes monolingual approaches that treat document written in different languages in a monolingual fashion. The intuition is that named entities (for example, ``Obama'') and cognate words (for example, ``tsunami'') are written in the same or similar fashion in many languages. For example, the Cross-Language Character n-Gram Model (CL-CNG)~\cite{plagiarism} represents documents as bags of character $n$-grams. Another approach is to use language dependent keyword lists based on cognate words ~\cite{pouliquen2008story}. These approaches may be suitable for comparing documents written in languages that share a writing system, which does not apply to the case of global news tracking.

@@

\section{Cross-lingual Document Similarity}
\label{sec:crosslingual}
Document similarity is an important component in techniques from text mining and natural language processing. Many techniques use the similarity as a black box, e.g., a kernel in Support Vector Machines. Comparison of documents (or other types of text snippets) in a monolingual setting is a well-studied problem in the field of information retrieval ~\cite{Salton88term-weightingapproaches}. We first formally introduce the problem followed by a description of  our approach.

@@

\noindent\textbf{Document representation.}
The standard vector space model~\cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to a word or a phrase in a fixed vocabulary. Formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.

@@

\noindent\textbf {Mono-lingual similarity.}
A common way of computing similarity between documents is \emph{cosine similarity},
$$sim(d_1, d_2) = \frac{\langle \phi(d_1), \phi(d_2)\rangle}{\|\phi(d_1)\| \|\phi(d_2)\|},$$
where $\langle \cdot,\cdot \rangle$ and $\|\cdot\|$ are standard inner product and Euclidean norm. When dealing with two or more languages, one could ignore the language information
and build a vector space using the union of tokens over the languages. A cosine similarity function in such a space can be useful to some extent, for example ``Internet'' or ``Obama'' may appear both in Spanish and English texts and the presence of such terms in both an English and a Spanish document would contribute to their similarity. In general however, large parts of vocabularies may not intersect. This means that given a language pair, many words in both languages cannot contribute to the similarity score. Such cases can make the similarity function very insensitive to the data.

\noindent\textbf {Cross-lingual similarity.}
Processing a multilingual dataset results in several vector spaces with varying dimensionality, one for each language. The dimensionality of the vector space corresponding to the $i$-th language is denoted by $n_i$ and the vector space model mapping is denoted by $\phi_i : \text{text} \rightarrow \RR^{n_i}$.
The similarity between documents in language $i$ and language $j$ is defined as a bilinear operator represented as a matrix $S_{i,j} \in \RR^{n_i \times n_j}$:
$$sim_{i,j}(d_1, d_2) = \frac{ \langle \phi_i (d_1), S_{i,j} \phi_j (d_2) \rangle }{\|\phi_i(d_1)\| \|\phi_j(d_2)\|},$$
where $d_1$ and $d_2$ are documents written in the $i$-th and $j$-th language respectively. If the maximal singular value of $S_{i,j}$ is bounded by $1$, then the similarity scores will lie on the interval $[-1, 1]$. We will provide an overview of the models in Section \ref{sec:models} and then introduce additional notation in \ref{sec:notation}. Starting with Section \ref{sec:kmeans} and ending with Section \ref{sec:hublang} we will describe some approaches to compute $S_{i,j}$ given training data.

\subsection{Cross-Lingual Models}\label{sec:models}
In this section, we will describe several approaches to the problem of computing the multilingual similarities introduced in Section~\ref{sec:tfidf}. We present four approaches:
a simple approach based on $k$-means clustering in Section~\ref{sec:kmeans}, a standard approach based on singular value decomposition in Section~\ref{sec:LSI}, a related
approach called Canonical Correlation Analysis (CCA) in Section~\ref{sec:CCA} and finally a new method, which is an extension of CCA to more than two languages in Section~\ref{sec:hublang}.
%
CCA can be used to find correlated patterns for a pair of languages, whereas the extended method optimizes a
Sum of Squared Correlations (SSCOR) between several language pairs, which was introduced in~\cite{Kettenring}. The SSCOR problem is difficult to solve in our setting (hundreds of thousands of features, hundreds of thousands of examples). To tackle this, we propose a method which consists of two ingredients.
 The first one is based on an observation that certain datasets (such as Wikipedia) are biased towards one language (English for Wikipedia), which can be exploited
 to reformulate a difficult optimization problem as an eigenvector problem. The second ingredient is dimensionality reduction using CL-LSI, which
 makes the eigenvector problem computationally and numerically tractable.

@@

The cross-lingual similarity models presented in this paper are based on comparable corpora. A \emph{comparable corpus} is a collection of documents in multiple languages, with alignment between documents that are of the same topic, or even a rough translation of each other. Wikipedia is an example of a comparable corpus, where a specific entry can be described in multiple languages (e.g., ``Berlin" is currently described in 222 languages). News articles represent another example, where the same event can be described by newspapers in several languages.

More formally, a \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a tuple of $m$ documents on the same topic (comparable), where $u_i$ is the document written in language $i$. Note that an individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least \textbf{two nonempty documents}. This means that in our analysis we discard strictly monolingual documents for which no cross-lingual information is available. A comparable corpus $D = {d_1, \ldots, d_s}$ is a collection of $s$ multilingual documents. By using the vector space model, we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times s}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$. Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$ and the matrices respect the document alignment - the vector $X_i^\ell$ corresponds to the TFIDF vector of the $i$-th component of multilingual document $d_\ell$. We use $N$ to denote the total row dimension of $X$, i.e., $N:= \sum_{i=1}^m n_i$. See Figure~\ref{fig:stacked_matrices} for an illustration of the introduced notation.

\begin{figure}[tbp]
\centering
\includegraphics[width=9cm]{stacked_matrices1-crop.pdf}
\caption{\label{fig:stacked_matrices} Multilingual corpora and their matrix representations using the vector space model.}
\end{figure}

We will now describe four models to cross-lingual similarity computation in the next sub-sections.
\subsection{$k$-means}\label{sec:kmeans}

The $k$-means algorithm is perhaps the most well-known and widely-used clustering algorithm. Here, we present its application
to compute cross-lingual similarities. The idea is based on concatenating the corpus matrices, running standard $k$-means clustering to obtain the matrix of centroids, ``reversing" the concatenation step to obtain a set of aligned bases, which are finally used to compute cross-lingual similarities. See Figure~\ref{fig:kmeans} for overview of the procedure. The left side of Figure~\ref{fig:kmeans} illustrates the decomposition and the right side summarizes the coordinate change.

@@

We then run the $k$-means algorithm~\cite{kmeans} and obtain a centroid matrix $C \in \RR^{N \times k}$, where the $k$ columns represent centroid vectors. The centroid matrix can be split vertically into $m$ blocks: $$C = [C_1^T \cdots C_m^T]^T,$$ according to the number of dimensions of each language, i.e., $C_i \in \RR^{n_i \times k}$.
%
To reiterate, the matrices $C_i$ are computed using a multilingual corpus matrix $X$ (based on Wikipedia for example).

@@

The resulting matrix for similarity computation between language $i$ and language $j$ is defined up to a scaling factor as:
$$C_i(C_i^T C_i)^{-1} (C_j^T C_j)^{-1} C_j.$$

The matrix is a result of mapping documents in a language independent space using pseudo-inverses of the centroid matrices $P_i = (C_i^T C_i)^{-1} C_i$ and then comparing them using the standard inner product, which results in the matrix $P_i^T P_j$. For the sake of presentation, we assumed that the centroid vectors are linearly independent. (An independent subspace could be obtained using an additional Gram-Schmidt step~\cite{golub} on the matrix $C$, if this was not the case.)

\subsection{Cross-Lingual Latent Semantic Indexing}\label{sec:LSI}

The second approach we consider is Cross-Lingual Latent Semantic Indexing (CL-LSI)~\cite{cl_lsi} which is a variant of LSI~\shortcite{lsi} for more than one language. The approach is very similar to $k$-means, where we first concatenate the corpus matrices, compute a decomposition, which in case of CL-LSI is a truncated Singular Value Decomposition (SVD), decouple the
 column space matrix and use the blocks to compute linear maps to a common vector space, where standard cosine similarity is used to compare documents.
